Revisiting the hydrodynamic modulation of short surface waves by longer waves

### 2025-03-11 paper.tex
Finishing touches

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the wave crest and action conservation laws and derive
higher-order steady solutions for the change of short-wave wavenumber, action,
and gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
The short-wave steepness modulation is attributed to be 5/8 due to the
wavenumber, 1/4 due to the wave action, and 1/8 due to the effective gravity.
Examining the homogeneity and stationarity requirements of the wave action
conservation equation reveals that stationarity is more easily violated than
homogeneity, and that the short-wave action is strongly conserved ($>$99\%) only
for very small-amplitude long waves, and weakly conserved ($>$90\%) for steep
long waves.
We further examine the results of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The proposed steady solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx$0.2.
Except for a subset of initial conditions, the solutions to the fully nonlinear
crest-action conservation equations are mostly steady in the reference frame of
the long waves.
\end{abstract}

@@

Why is it, then, that the solution for short-wave modulation is not only
quantitatively, but also qualitatively, different depending on the short-wave
initial conditions?
It may at first seem that there is something special about the uniform initial
conditions that makes its solutions grow unsteadily and indefinitely.
\citet{peureux2021unsteady} described this scenario as "...the sudden appearance
of a long wave perturbation in the middle of a homogeneous short wave field".
However, it is not obvious whether such a scenario is realistic as waves tend to
disperse into groups, and the time scales of incoming swell fronts are much
longer than those of short-wave generation by wind.
To test whether the sudden onset of long-wave forcing causes the unsteady growth,
we repeat the numerical simulation of Fig.
\ref{fig:modulation_3panel_infinite}, except for a linear ramp applied to $a_L$
(and consequently, the other long-wave properties including the orbital
velocities and the local steepness) for the first five periods:
$a_L(t) = min(a_L, a_L t / \left(5 T_L\right))$, where $T_L$ is the long-wave period.
This result is shown in Fig. \ref{fig:modulation_3panel_ramp}.
The key distinction that arises with the introduction of the ramp is that the
modulation pattern of short waves is now locked-in to the long-wave crests,
and their indefinite steepening vanishes.
The stabilisation effect of the long-wave ramp is more readily assessed in
Fig. \ref{fig:unsteady_growth_timeseries}, which shows the maximum short-wave
steepness modulation as function of time over 30 long-wave periods, for the
infinite long-wave train case and the case of applying a linear ramp.
Comparing the wave action propagation ($C_g \partial N / \partial x$) and
inhomogeneity ($N \partial C_g / \partial x$) tendencies reveals that
after only one long-wave period, the propagation and inhomogeneity tendencies
are unbalanced and amplify the steepening of short waves
(Fig. \ref{fig:inhomogeneity_tendencies}).
\citet{peureux2021unsteady} correctly pointed out that the unsteady growth of
short-wave steepness occurs due to the resonance of the inhomogeneity
tendency with the short-wave action perturbation.
However, in the case of a gentle ramp-up of the long-wave forcing, the
propagation and inhomogeneity tendencies mostly balance each other and allow
the short-wave steepening to lock-in to the long-wave crests.

@@

We now proceed to examine the numerical solutions of short-wave modulation by
long waves whose amplitude varies with time.
The amplitude variation aims to mimic the behaviour of long-wave groups, a
phenomenon that is ubiquitous in the ocean due to the dispersive nature of
surface gravity waves, as well as due to the fact that wind-generated waves
evolve to have a broad-banded spectra.
For simplicity of implementation and interpretation, 
the wave group here is simulated as a non-dispersive train of long waves whose
amplitude scales with $\sin[\pi t / (n T_L)]$, where $n$ is the number of
long waves in the group.
The elevation of the long waves thus gradually ramps up from rest to $a_L$ at
$t = n T_L / 2$, after which it gradually decreases toward zero.
In this simulation, $n = 10$ and $\varepsilon_L = 0.1$ as in the previous two
simulations.
The modulations of short-wave wavenumber, amplitude, and steepness in response
to such long-wave group is shown in Fig. \ref{fig:modulation_3panel_groups}.
Like in the case of a linear ramp, the modulation pattern here is locked in to
the long-wave crest and peaks at approximately 1.2 for short-wave steepness of 0.1.
After the peak of the long-wave group passes, the short-wave modulation recedes
back toward the resting state.

@@

Although in this paper we only explored the solutions for $k_L = 1$ and $k=10$
the results are not sensitive to different values of $k_L$, provided the same
long-wave steepness $\varepsilon_L = a_L k_L$.
Numerical simulations with $k/k_L = 10$ and $k/k_L = 100$ show little
sensitivity to the ratio $k/k_L$ (not shown), consistent with the results of
\citet{longuet1987propagation}.
Specifically, the short-wave steepness modulation is 1\% larger at
$\varepsilon_L = 0.27$ for $k/k_L = 10$ compared to $k/k_L = 100$, and 4\%
larger at $\varepsilon_L = 0.4$.
This suggests that as long as $k/k_L \gg 1$, the modulation amplitude is
largely independent of the ratio $k/k_L$.
These results can be reproduced with the companion numerical model.
Also not considered here are the short-wave trains that are oblique to the
longer waves, which is relatively straightforward to include
\citep{peureux2021unsteady}.

@@

In this paper we revisited the problem of short-wave modulation by long waves
from a wave action conservation perspective.
Using the linearised wave crest and action conservation equations coupled with
nonlinear solutions for the effective gravity, we derived steady analytical
solutions for the modulation of short wave wavenumber, amplitude, steepness,
intrinsic frequency, and phase speed.
The latter two were previously examined in the context of short-wave
modulation \citep{longuet1962resonant,longuet1962phase}, and may be important
for correctly calculating the wind input into short waves, which is proportional
to the wind speed relative to the wave celerity.
The approximate steady solutions yield higher wavenumber and amplitude
modulation than that of \citet{longuet1960changes}, and somewhat lower than 
the numerical solutions of \citet{longuet1987propagation} and
\citet{zhang1990evolution}.
This is mainly due to the analytical solution requiring the crest and action
conservation equations to be linearised, and the nonlinearity arising due
to the evaluation of short waves on the surface of the long wave (as opposed to
the mean water level) and due to nonlinear effective gravity.
In the steady solutions, the short-wave steepness modulation is dominated by
the wavenumber modulation (5/8), followed by the wave action modulation (1/4),
and is least affected by the effective gravity modulation (1/8).

The numerical solutions of the full wave crest and action conservation equations
are consistent with the analytical solutions for $\varepsilon_L \lesssim 0.2$,
and are similar in amplitude to the numerical solutions of
\citet{longuet1987propagation} and \citet{zhang1990evolution}.
The simulations also reveal that the unsteady growth of short-wave steepness
reported by \citet{peureux2021unsteady} is due to the sudden appearance of
long waves in a homogeneous short-wave field, a scenario that is unlikely to
occur in the open ocean.
A more likely condition such as that of long-wave groups, are only mildly
destabilising for the short waves.
These results suggest that for mildly-to-moderately steep short waves,
the analytical steady solutions presented here are sufficient to describe the
modulation of short waves by long waves.
It may be worthwhile to revisit a more systematic inclusion of the hydrodynamic
modulation effects in phase-averaged spectral wave models, beyond the simple
cumulative mean squared slope for wave dissipation, but also for modulating
the dispersion and wind input into short waves.

>>>>

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the wave crest and action conservation laws and derive
higher-order steady solutions for the change of short-wave wavenumber, action,
and gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
The short-wave steepness modulation is attributed to be 5/8 due to the
wavenumber, 1/4 due to the wave action, and 1/8 due to the effective gravity.
Examining the homogeneity and stationarity requirements for the conservation of
wave action reveals that stationarity is a stronger requirement and is
generally not satisfied for very steep long waves.
We further examine the results of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The proposed steady solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx$0.2.
Except for a subset of initial conditions, the solutions to the non-linearised
crest-action conservation equations are mostly steady in the reference frame of
the long waves.
\end{abstract}

@@

Why is it, then, that the solution for short-wave modulation is not only
quantitatively, but also qualitatively, different depending on the short-wave
initial conditions?
It may at first seem that there is something special about the uniform initial
conditions that makes its solutions grow unsteadily and indefinitely.
\citet{peureux2021unsteady} described this scenario as "...the sudden appearance
of a long wave perturbation in the middle of a homogeneous short wave field".
However, it is not obvious whether such a scenario is common as waves tend to
disperse into groups, and the time scales of incoming swell fronts are much
longer than those of short-wave generation by wind.
Exceptional situations with such a sudden onset of swell may occur could be
around complex coastlines or in shadows behind islands, where wind-generated
ripples may propagate from an otherwise calm sea state into the path of a
crossing swell.

To test whether the sudden onset of long-wave forcing causes the unsteady growth,
we repeat the numerical simulation of Fig.
\ref{fig:modulation_3panel_infinite}, except for a linear ramp applied to $a_L$
(and consequently, the other long-wave properties including the orbital
velocities and the local steepness) for the first five periods:
$a_L(t) = min(a_L, a_L t / \left(5 T_L\right))$, where $T_L$ is the long-wave period.
This result is shown in Fig. \ref{fig:modulation_3panel_ramp}.
The key distinction that arises with the introduction of the ramp is that the
modulation pattern of short waves is now locked-in to the long-wave crests,
and their indefinite steepening vanishes.
The stabilisation effect of the long-wave ramp is more readily assessed in
Fig. \ref{fig:unsteady_growth_timeseries}, which shows the maximum short-wave
steepness modulation as function of time over 30 long-wave periods, for the
infinite long-wave train case and the case of applying a linear ramp.
Comparing the wave action propagation ($C_g \partial N / \partial x$) and
inhomogeneity ($N \partial C_g / \partial x$) tendencies reveals that
after only one long-wave period, the propagation and inhomogeneity tendencies
are unbalanced and amplify the steepening of short waves
(Fig. \ref{fig:inhomogeneity_tendencies}).
\citet{peureux2021unsteady} correctly pointed out that the unsteady growth of
short-wave steepness occurs due to the resonance of the inhomogeneity
tendency with the short-wave action perturbation.
However, in the case of a gentle ramp-up of the long-wave forcing, the
propagation and inhomogeneity tendencies mostly balance each other and allow
the short-wave steepening to lock-in to the long-wave crests.

@@

We now proceed to examine the numerical solutions of short-wave modulation by
long waves whose amplitude varies with time.
The amplitude variation aims to mimic the behaviour of long-wave groups, a
phenomenon that is ubiquitous in the ocean due to the dispersive nature of
surface gravity waves, as well as due to the fact that wind-generated waves
evolve to have broad-banded spectra.
For simplicity of implementation and interpretation, 
the wave group here is simulated as a non-dispersive train of long waves whose
amplitude scales with $\sin[\pi t / (n T_L)]$, where $n$ is the number of
long waves in the group.
The elevation of the long waves thus gradually ramps up from rest to $a_L$ at
$t = n T_L / 2$, after which it gradually decreases toward zero.
In this simulation, $n = 10$ and $\varepsilon_L = 0.1$ as in the previous two
simulations.
The modulations of short-wave wavenumber, amplitude, and steepness in response
to such long-wave group is shown in Fig. \ref{fig:modulation_3panel_groups}.
Like in the case of a linear ramp, the modulation pattern here is locked in to
the long-wave crest and peaks at approximately 1.2 for short-wave steepness of 0.1.
After the peak of the long-wave group passes, the short-wave modulation recedes
back toward the resting state.

@@

Although in this paper we only explored the solutions for $k_L = 1$ and $k=10$
the results are not sensitive to different values of $k_L$, provided the same
long-wave steepness $\varepsilon_L = a_L k_L$.
Numerical simulations with $k/k_L = 10$ and $k/k_L = 100$ show little
sensitivity to the ratio $k/k_L$ (not shown), consistent with the results of
\citet{longuet1987propagation}.
Specifically, the short-wave steepness modulation is 1\% larger at
$\varepsilon_L = 0.27$ for $k/k_L = 10$ compared to $k/k_L = 100$, and 4\%
larger at $\varepsilon_L = 0.4$.
This suggests that as long as $k/k_L \gg 1$, the modulation amplitude is
largely independent of the ratio $k/k_L$.
These results can be reproduced with the companion numerical model.
Note, however, according to \S\ref{sec:wave_action_conservation}, that the
wave action conservation requirements are more easily satisfied for larger
$k/k_L$.
Also not considered here are the short-wave trains that are oblique to the
longer waves, which is relatively straightforward to include
\citep{peureux2021unsteady}.

@@

In this paper we revisited the problem of short-wave modulation by long waves
from a wave action conservation perspective.
Using the linearised wave crest and action conservation equations coupled with
nonlinear solutions for the effective gravity, we derived steady analytical
solutions for the modulation of short wave wavenumber, amplitude, steepness,
intrinsic frequency, and phase speed.
The latter two were previously examined in the context of short-wave
modulation \citep{longuet1962resonant,longuet1962phase}, and may be important
for correctly calculating the wind input into short waves, which is proportional
to the wind speed relative to the wave celerity.
The approximate steady solutions yield higher wavenumber and amplitude
modulation than that of \citet{longuet1960changes}, and somewhat lower than 
the numerical solutions of \citet{longuet1987propagation} and
\citet{zhang1990evolution}.
This is mainly due to the analytical solution requiring the crest and action
conservation equations to be linearised, and the nonlinearity arising due
to the evaluation of short waves on the surface of the long wave (as opposed to
the mean water level) and due to nonlinear effective gravity.
In the steady solutions, the short-wave steepness modulation is dominated by
the wavenumber modulation (5/8), followed by the wave action modulation (1/4),
and is least affected by the effective gravity modulation (1/8).
The homogeneity and stationarity criteria for wave action conservation are
evaluated for both the analytical and numerical solutions.
The stationarity is a stronger criterion than homogeneity, and is found to be
strongly satisfied (> 99\% stationary) for very gently sloped waves and weakly
satisfied (> 90\% stationary) for steeper waves.

The numerical solutions of the full wave crest and action conservation equations
are consistent with the analytical solutions for $\varepsilon_L \lesssim 0.2$,
and are similar in amplitude to the numerical solutions of
\citet{longuet1987propagation} and \citet{zhang1990evolution}.
The simulations also reveal that the unsteady growth of short-wave steepness
reported by \citet{peureux2021unsteady} is due to the sudden appearance of
long waves in a homogeneous short-wave field, a scenario that is unlikely to
be common in the open ocean.
A more likely condition such as that of long-wave groups, is only mildly
destabilising for the short waves.
These results suggest that for the long-wave steepness $\varepsilon_L \lesssim 0.2$,
the analytical steady solutions presented here are sufficient to describe the
modulation of short waves by long waves.
For moderately steep long waves ($0.2 \lesssim \varepsilon_L \lesssim 0.3$),
numerical solutions of non-linearised crest and action conservation equations
are necessary to capture the short-wave steepness modulation.
For steeper long waves ($\varepsilon_L > 0.3$), the short-wave properties can
no longer be considered stationary \citep{bretherton1968wavetrains}, and the
wave action conservation equation is not valid in this context.
It may be worthwhile to revisit a more systematic inclusion of the hydrodynamic
modulation effects in phase-averaged spectral wave models, beyond the simple
cumulative mean squared slope for wave dissipation, but also for modulating
the dispersion and wind input into short waves.
